---
title: "Short Pres - Doubles for Dollars"
author: "Sam Turner"
format:
  html: 
    self-contained: true
editor: visual
execute: 
  warning: false
---

# Load Libraries

```{r}
library(Lahman)
library(tidyverse)
library(broom)
library(fixest)
```

# Load Datasets & Filter

```{r}
Teams <- Lahman::Teams

Teams1985 <- Teams %>% 
  filter(yearID >= 1985 & yearID <= 2016) %>% 
  mutate(X1B = H - X2B - X3B - HR, .after = H) %>% 
  mutate(OB = BB + HBP, .after = HBP)

#Put this after because ">=" no work with factor variables
Teams1985$yearID <- as.factor(Teams1985$yearID)

Teams1985 %>%
  select(yearID, teamID, R, AB, H, X1B, X2B, X3B, HR, BB, SO, SB, CS, HBP, OB, SF) %>% 
  glimpse()
```

# Data Description

| Variable | Min                    | Q1                                     | Median                    | Q3                                     | Max                    | Mean                                     | SD                                     |
|---------|---------|---------|---------|---------|---------|---------|---------|
| X1B      | `r min(Teams1985$X1B)` | `r quantile(Teams1985$X1B, probs=.25)` | `r median(Teams1985$X1B)` | `r quantile(Teams1985$X1B, probs=.75)` | `r max(Teams1985$X1B)` | `r round(mean(Teams1985$X1B), digits=2)` | `r round(sd(Teams1985$X1B), digits=2)` |
| X2B      | `r min(Teams1985$X2B)` | `r quantile(Teams1985$X2B, probs=.25)` | `r median(Teams1985$X2B)` | `r quantile(Teams1985$X2B, probs=.75)` | `r max(Teams1985$X2B)` | `r round(mean(Teams1985$X2B), digits=2)` | `r round(sd(Teams1985$X2B), digits=2)` |
| X3B      | `r min(Teams1985$X3B)` | `r quantile(Teams1985$X3B, probs=.25)` | `r median(Teams1985$X3B)` | `r quantile(Teams1985$X3B, probs=.75)` | `r max(Teams1985$X3B)` | `r round(mean(Teams1985$X3B), digits=2)` | `r round(sd(Teams1985$X3B), digits=2)` |
| HR       | `r min(Teams1985$HR)`  | `r quantile(Teams1985$HR, probs=.25)`  | `r median(Teams1985$HR)`  | `r quantile(Teams1985$HR, probs=.75)`  | `r max(Teams1985$HR)`  | `r round(mean(Teams1985$HR), digits=2)`  | `r round(sd(Teams1985$HR), digits=2)`  |
| OB       | `r min(Teams1985$OB)`  | `r quantile(Teams1985$OB, probs=.25)`  | `r median(Teams1985$OB)`  | `r quantile(Teams1985$OB, probs=.75)`  | `r max(Teams1985$OB)`  | `r round(mean(Teams1985$OB), digits=2)`  | `r round(sd(Teams1985$OB), digits=2)`  |
| SB       | `r min(Teams1985$SB)`  | `r quantile(Teams1985$SB, probs=.25)`  | `r median(Teams1985$SB)`  | `r quantile(Teams1985$SB, probs=.75)`  | `r max(Teams1985$SB)`  | `r round(mean(Teams1985$SB), digits=2)`  | `r round(sd(Teams1985$SB), digits=2)`  |
| SF       | `r min(Teams1985$SF)`  | `r quantile(Teams1985$SF, probs=.25)`  | `r median(Teams1985$SF)`  | `r quantile(Teams1985$SF, probs=.75)`  | `r max(Teams1985$SF)`  | `r round(mean(Teams1985$SF), digits=2)`  | `r round(sd(Teams1985$SF), digits=2)`  |

# Fixed Effects Model

```{r}
feols(R ~ X1B + X2B + X3B + HR + OB + SF | yearID + teamID, data = Teams1985) %>% 
  tidy()
```

## Formula

Current formula:

$$
\widehat{xRC}=.47X1B+.69X2B+1.22X3B+1.41HR+.32OB+.85SF
$$
